---
title: "HW 2 Template"
author: "Fill Name in"
date: "Due 9/14/2022"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


---

### HW 2 - Part 1: **`glimpse`** and **`unique`**

#### Chunk 2: Examining the **`diamonds`** Data

- This chunk reviews the **`glimpse`** and **`unique`** commands from Week 1

- You are encouraged to add your own notes here and/or your own comments in the chunk below so that you understand the code and what it does.

```{r examining the diamonds data}

# save the diamonds dataset to your Global Environment
my_diamonds <- diamonds

# show data summary
# without piping
glimpse(my_diamonds)

# with piping
my_diamonds |>
  glimpse()

# examine ordinal variables using unique
# without piping
unique(my_diamonds$cut)

# with piping
# pull command selects a single variable from a dataset
my_diamonds |> pull(cut) |> unique()

# choose one of these methods to examine:


# ordered levels of clarity


# ordered levels of color


```

---

### HW2 - Part 2: **`select`**

#### Chunk 3: Selecting variables in a dataset

- This chunk demonstrates using the **`select`** command to select variables.

- You are encouraged to add your own notes here and/or your own comments in the chunk below so that you understand the code and what it does.

```{r selecting variables in a dataset}

# creates a new dataset, my_diamonds1
# selects variables we want to keep
# select can also re-order variables (price is now first)
my_diamonds1 <- my_diamonds |>
  select(price, carat, cut, color, clarity, depth, table)

# same command, but shorter
# used : to specify consecutive variables
my_diamonds1 <- my_diamonds |>
  select(price, carat:table)

# creates a new dataset, my_diamonds2
# identical to above but we use ! to exclude variables
my_diamonds2 <- my_diamonds |>
  select(!c(x,y,z))

# create my_diamonds3 using select to keep these variables only: 
# price, carat, color, cut, clarity



# create my_diamonds4 using select to EXCLUDE these variables 
# depth, table, x, y, z



```

---

### HW2 - Part 3: **`slice`**

#### Chunk 4: Selecting rows by row number

##### Editors: Author(s) who edited Chunk 4

- This chunk demonstrates using the **`slice`** command to select rows.

- You are encouraged to add your own notes here and/or your own comments in the chunk below so that you understand the code and what it does.

```{r slicing data by row number}

my_diamonds1 <- my_diamonds |>
  select(price, carat, cut, color, clarity, depth, table) |>
  slice(5001:35000)

my_diamonds2 <- my_diamonds |>
  select(price, carat:table) |>
  slice(1, 200, 2301:3000, 29000:32000)

# Use the code from Chunk 3  that you wrote to create my_diamonds3
# add on to that code to limit my_diamonds3 to rows:
# 1001 through 30000 and 45001 through 50000



```

---

### HW 2 - Part 4: **`filter`** and **`summary`**

#### Chunk 5: Filtering data by value and summarizing

- This chunk demonstrates using the **`filter`** command to select rows by variables values and using **`summary`**

- You are encouraged to add your own notes here and/or your own comments in the chunk below so that you understand the code and what it does.

```{r filtering data by value and summarizing}

# filtering my_diamonds1 to 
# carat weight greater than or equal to 2 (carat >=2)
# cut is Premium or Ideal (cut %in% c("Premium", "Ideal"))
my_diamonds1 <- my_diamonds |>
  select(price, carat:table) |>
  slice(5001:35000) |>
  filter(carat >= 2) |>
  filter(cut %in% c("Premium", "Ideal"))

# same filtering tasks as above
# !cut used to exclude cuts Fair, Good, Very Good
my_diamonds1 <- my_diamonds |>
  select(price, carat:table) |>
  slice(5001:35000) |>
  filter(carat >= 2) |>
  filter(!cut %in% c("Fair", "Good", "Very Good"))

# same filtering tasks as above
# both filtering tasks combined using and operator &
my_diamonds1 <- my_diamonds |>
  select(price, carat:table) |>
  slice(5001:35000) |>
  filter(carat >= 2 & cut %in% c("Premium", "Ideal")) 

# summary of color in my_diamonds1
summary(my_diamonds1$color)

# same summary with piping
my_diamonds1 |> pull(color) |> summary()

# Use any one of the examples above to filter my_diamonds3 to:
# carat weight greater than or equal to 1.25
# cut is Very Good, Premium, or Ideal 


# summarize clarity in the my_diamonds3 dataset (number of obs. in each level)
# use the summary command, with or without piping
# above I show how I summarized the color variable


```

---


### HW 2 - Part 5: Creating plots with **`ggplot`**

#### Creating Basic Plots

- In Part 5, you will create a chunk (Ctrl/Cmd + Alt + i) and copy the R code for Part 5 from **HW2_Instructions.html** into the chunk you create.

- The provided R code demonstrates using 

   - **`ggplot`** to make some basic plots
  
   -  **`grid.arrange`** to present multiple plots in a grid or column.
  
- You are encouraged to add your own notes here and/or your own comments in the chunk you create so that you understand the code and what it does.

