---
title: "HW 2 Solution"
author: "Penelope Pooler"
date: "Due 1/31/2024"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=T}

# this line specifies options for default options for all R Chunks
knitr::opts_chunk$set(echo=T, highlight=T)

# suppress scientific notation
options(scipen=100)

# install helper package (pacman) if needed
if (!require("pacman")) install.packages("pacman", repos = "http://lib.stat.cmu.edu/R/CRAN/")

# install and load required packages
# pacman should be first package in parentheses and then list others
pacman::p_load(pacman, tidyverse, gridExtra, magrittr)

# verify packages (comment out in finished documents)
p_loaded()


```


---

### HW 2 - Part 1: **`glimpse`** and **`unique`**

#### Chunk 2: Examining the **`diamonds`** Data

- This chunk reviews the **`glimpse`** and **`unique`** commands from Week 1

- You are encouraged to add your own notes here and/or your own comments in the chunk below so that you understand the code and what it does.

```{r examining the diamonds data}

# save the diamonds dataset to your Global Environment
my_diamonds <- diamonds

# show data summary
# without piping
glimpse(my_diamonds)

# with piping
my_diamonds |>
  glimpse(width = 50)

# examine ordinal variables using unique
# without piping
unique(my_diamonds$cut)

# with piping
# pull command selects a single variable from a dataset
my_diamonds |> pull(cut) |> unique()

# choose one of these methods to examine:

# ordered levels of clarity
my_diamonds |> pull(clarity) |> unique()

# ordered levels of color
unique(my_diamonds$color)


```

---

### HW2 - Part 2: **`select`**

#### Chunk 3: Selecting variables in a dataset

- This chunk demonstrates using the **`select`** command to select variables.

- You are encouraged to add your own notes here and/or your own comments in the chunk below so that you understand the code and what it does.

```{r selecting variables in a dataset}

# creates a new dataset, my_diamonds1
# selects variables we want to keep
# select can also re-order variables (price is now first)
my_diamonds1 <- my_diamonds |>
  select(price, carat, cut, color, clarity, depth, table)

# same command, but shorter
# used : to specify consecutive variables
my_diamonds1 <- my_diamonds |>
  select(price, carat:table)

# creates a new dataset, my_diamonds2
# identical to above but we use ! to exclude variables
my_diamonds2 <- my_diamonds |>
  select(!c(x,y,z))

# create my_diamonds3 using select to keep these variables only: 
# price, carat, color, cut, clarity

my_diamonds3 <- my_diamonds |>
  select(price, carat, cut, color, clarity)

# create my_diamonds4 using select to EXCLUDE these variables 
# depth, table, x, y, z

my_diamonds4 <- my_diamonds |>
  select(!c(x,y,z, depth, table))


```

---

### HW2 - Part 3: **`slice`**

#### Chunk 4: Selecting rows by row number

##### Editors: Author(s) who edited Chunk 4

- This chunk demonstrates using the **`slice`** command to select rows.

- You are encouraged to add your own notes here and/or your own comments in the chunk below so that you understand the code and what it does.

```{r slicing data by row number}

my_diamonds1 <- my_diamonds |>
  select(price, carat, cut, color, clarity, depth, table) |>
  slice(5001:35000)

my_diamonds2 <- my_diamonds |>
  select(price, carat:table) |>
  slice(1, 200, 2301:3000, 29000:32000)

# Use the code from Chunk 3  that you wrote to create my_diamonds3
# add on to that code to limit my_diamonds3 to rows:
# 1001 through 30000 and 45001 through 50000

my_diamonds3 <- my_diamonds |>
  select(price, carat, cut, color, clarity) |>
  slice(1001:30000, 45001:50000)

```

---

### HW 2 - Part 4: **`filter`** and **`summary`**

#### Chunk 5: Filtering data by value and summarizing

- This chunk demonstrates using the **`filter`** command to select rows by variables values and using **`summary`**

- You are encouraged to add your own notes here and/or your own comments in the chunk below so that you understand the code and what it does.

```{r filtering data by value and summarizing}

# filtering my_diamonds1 to 
# carat weight greater than or equal to 2 (carat >=2)
# cut is Premium or Ideal (cut %in% c("Premium", "Ideal"))
my_diamonds1 <- my_diamonds |>
  select(price, carat:table) |>
  slice(5001:35000) |>
  filter(carat >= 2) |>
  filter(cut %in% c("Premium", "Ideal"))

# same filtering tasks as above
# !cut used to exclude cuts Fair, Good, Very Good
my_diamonds1 <- my_diamonds |>
  select(price, carat:table) |>
  slice(5001:35000) |>
  filter(carat >= 2) |>
  filter(!cut %in% c("Fair", "Good", "Very Good"))

# same filtering tasks as above
# both filtering tasks combined using and operator &
my_diamonds1 <- my_diamonds |>
  select(price, carat:table) |>
  slice(5001:35000) |>
  filter(carat >= 2 & cut %in% c("Premium", "Ideal")) 

# summary of color in my_diamonds1
summary(my_diamonds1$color)

# same summary with piping
my_diamonds1 |> pull(color) |> summary()

# Use any one of the examples above to filter my_diamonds3 to:
# carat weight greater than or equal to 1.25
# cut is Very Good, Premium, or Ideal 

my_diamonds3 <- my_diamonds |>
  select(price, carat, cut, color, clarity) |>
  slice(1001:30000, 45001:50000) |>
  filter(carat >= 1.25) |>
  filter(cut %in% c("Very Good", "Premium", "Ideal"))


# summarize clarity in the my_diamonds3 dataset (number of obs. in each level)
# use the summary command, with or without piping
# above I show how I summarized the color variable

my_diamonds3 |> pull(clarity) |> summary()


```

---


### HW 2 - Part 5: Creating plots with **`ggplot`**

#### Creating Basic Plots

- In Part 5, you will create a chunk (Ctrl/Cmd + Alt + i) and copy the R code for Part 5 from **HW2_Instructions.html** into the chunk you create.

- The provided R code demonstrates using 

   - **`ggplot`** to make some basic plots
  
   -  **`grid.arrange`** to present multiple plots in a grid or column.
  
- You are encouraged to add your own notes here and/or your own comments in the chunk you create so that you understand the code and what it does.


```{r creating plots with ggplot}


#### scatter plots ####

# scatter_none is the most basic scatter plot of carat vs. price 
# no other variables are included
# to view this plot by itself (not required), enclose code in parentheses
scatter_none <- my_diamonds |>
   ggplot() +
   geom_point(aes(x=carat, y=price))


# scatter_clarity adds the option color=clarity to the aes (aesthetic)
# observations are color coded by diamond clarity level
# theme_classic() added to remove background

scatter_clarity <- my_diamonds |>
  ggplot() +
  geom_point(aes(x=carat, y=price, color=clarity)) +
  theme_classic()


# create plot named scatter_cut using the above code
# change color=clarity to color=cut

scatter_cut <- my_diamonds |>
  ggplot() +
  geom_point(aes(x=carat, y=price, color=cut)) +
  theme_classic()


# create plot named scatter_color using the above code
# change color=clarity to color=color

scatter_color <-  my_diamonds |>
  ggplot() +
  geom_point(aes(x=carat, y=price, color=color)) +
  theme_classic()



# plot all 4 plots above in a 2x2 grid and answer Blackboard Questions

grid.arrange(scatter_none, scatter_clarity, 
              scatter_color, scatter_cut, ncol=2)


#### boxplots ####

# below is a plot of grouped side-by_side boxplots 
# Within each cut category there is a separate boxplot for each color
# this is one good way to examine categorical data

# code is enclosed in parentheses 
# plot is saved as box_color AND is shown on the screen

(box_color <- my_diamonds3 |>
  ggplot() +
  geom_boxplot(aes(x=cut, y=price, fill=color))+ 
  theme_classic())


# create a plot of grouped side-by_side boxplots that show
# boxplots for each clarity category with each cut category

# same plot as above but change fill=color to fill=clarity
# name this plot box_clarity

(box_clarity <- my_diamonds3 |>
  ggplot() +
  geom_boxplot(aes(x=cut, y=price, fill=clarity))+ 
  theme_classic())


# plot these two sets of plots in a stacked column (ncol=1)

grid.arrange(box_color, box_clarity, ncol=1)



```



